---
title: "User Drinking Behaviors by Income Levels"
subtitle: "Case of Market in Finland"
author: "Min"
date: last-modified
date-format: "MMMM DD, YYYY"
execute:
  warning: false
  echo: false
  error: false  
format: 
  pdf: 
    toc: false
    geometry:
     - top=2cm
     - left=2cm
     - right=2cm
     - bottom=3cm
    fig-pos: 'h'
    fig-cap-location: bottom
    tbl-cap-location: bottom
    include-in-header:
    - text: |
        \usepackage{placeins}
editor_options: 
  chunk_output_type: inline
editor: 
  markdown: 
    wrap: 72
bibliography: references.bib
---

# Introduction & Research Background

## Data

ESS(European Social Survey) is a large-scale, general population survey
based on face-to-face and conducted in many European countries every two
years. Since 2001, the ESS measures people’s attitudes,beliefs, and
behavior patterns. The dataset I am going to analyze here is from ESS11
edition 2.0, published 21.11.24. The full dataset can be found
here:https://ess.sikt.no/en/datafile/242aaa39-3bbb-40f5-98bf-bfb1ce53d8ef?tab=0
and
"https://ess.sikt.no/en/datafile/242aaa39-3bbb-40f5-98bf-bfb1ce53d8ef/128?tab=0&elems=07f7e265-98eb-4ee6-b448-0e8d3b9e9bd3_6

## Research question

This report presents an analysis of drinking behavior patterns in
different income groups in Finland during 2023.

Alcohol plays different roles in today's life. It can be a way for fun,
for socialization or business, and also work as a mirror of personal
mental health, so that drinking behaviors can be very different from
person to person, more than just an impact of the country drinking
culture.

In this analysis, I am going to select Finland respondents data from
ESS11 and research how the drinking behaviors are different from people
in various income levels, including their drinking frequency, when they
prefer to drink(weekend or weekday), and if gender, life satisfaction
will have an impact or not.

# Theory & Hypotheses

Different income level affects drinking patterns.High-income-level
people usually spend time on business, tending to drive their alcohol
consumption increase.

According to Bourdieu’s (1984) theory of cultural distinction, social
economic status shapes consumption patterns. Higher-income individuals
usually engage frequently in social drinking settings, such as business
events or networking gatherings. However, as for alcohol consumption,
The Set Point Theory of Happiness ([@diener1999]) suggests that
happiness levels remain stable despite changes in external conditions
like income.That means drinking patterns of high-income people can be
different based on their sense of happiness.

As [@repke2022] suggests, the quantity of alcohol consumed in all 21
European countries, men consume almost twice as many units as women, so
gender also has an impact, so it further implies that under the same
income level, female consumes less alcohol than male.

-   **Hypothesis 1.1**:Higher income people will have higher drinking
    frequency, but it can be moderated by the level of happiness.

-   **Hypothesis 1.2**: Alcohol consumption from females is lower than
    males, in the condition of same income level.

It's pointed by the Job Demand-Resource Model (Demerouti et al., 2001)
that workplace demands influence stress levels and coping behaviors.
High-income people may have more chances to access alcohol on weekdays
at the level of financial avilablity.

-   **Hypothesis 2**: High-income people are more likely to drink on
    weekdays, comparing with low & middle-income people.

```{r setup}


library(haven)
library(survival)
library(modelsummary)
library(tidyverse)
library(dplyr)
library(socviz) 
library(modelr)
library(haven)
library(nnet)
library(broom)
library(gt)
library(gtsummary)
library(kableExtra)

#load data
setwd("/Users/a1-6/Desktop/CSS/1. Discrete choice modeling/DCM Assignment/Final assignment")

ess11 <- read_dta("ESS11.dta")  
#str(ess11)


#codebook
ess11_cbook <-   ess11 |> 
  summarise(across(everything(), \(x) attr(x, "label"))) |> 
  pivot_longer(cols = -1,
               names_to = "variable",
               values_to = "label") |> 
  select(-1)

```

```{r}
ess11_1 <- ess11 |> 
  filter(cntry == "FI") |>  
  select(cntry,idno, gndr,agea,edlvdfi,sclmeet, health,stflife,happy, hinctnta, alcfreq,alcwkdy,alcwknd)  |> 
  drop_na() |> 
  rename(
    "income" = hinctnta,        
    "female" = gndr,            
    "age" = agea                
  ) |>
  mutate(
    "female" = as.integer(female ==2),
    "happy" = as.integer(happy >= 8),
    "healthy" = as.integer(health <= 2))


# define 3 levels 
ess11_1 <- ess11_1 |> 
  mutate(alc_level = case_when(
    alcfreq >= 6 ~ 1, #"alomost no alc",
    alcfreq%in% 4:5 ~ 2, #"low alc",
    alcfreq <= 3 ~ 3 #"frequent alc"
  ),
  alc_level = factor(alc_level, 
                       levels = c(1, 2, 3),
                       labels = c("Almost No", "Low", "Frequent")),
  income_level = case_when(
   income <= 4 ~ 1,#"Low Income",
  income %in% 5:7 ~ 2,#"Mid Income",
  income >= 8 ~ 3 #"High Income"   
  ),
  income_level = factor(income_level, 
                       levels = c(1, 2, 3),
                       labels = c("Low Income", "Middle Income", "High Income")))

```

# Descriptive statistics

## Variable

According to the research questions, here're some variables being
selected from ESS11.

-   "Alcfreq" is used to measure "how often to drink alcohol". It's an
    ordinal varaible ranging from 1-7. (01 Every day;02 Several times a
    week;03 Once a week;04 2-3 times a month;05 Once a month;06 Less
    than once a month;07 Never). To differentiate their alcohol
    comsumption frequency, this variable will be divided into 3
    different levels: "High Alc"(at least once per week), "Low Alc(1-3
    times per month)" and "Almost No(under "less than once a month")".
    From @tbl-1, we find 3 different drink frequency groups are
    distributed in a balanced way, so it has enough data for future
    models.

-   "hinctnta" in ESS indicates "Household's total net income, all
    sources" and is used as the measurement of income. It has 10 deciles
    and the median is the fifth decile. As The 80-20 Rule, the top 20%
    will be catagorized into "High income". And the people who don't
    reach the mean value will be catagorized into "Low income", so the
    others will be grouped as ""Middle income".

-   "happy" is "How happy are you", and is measured by 10 levels from
    1("extremely unhappy")-10(extremely happy). In this analysis, I am
    define the people rating above and equal to 8 as "Happy" and set the
    dummy variable as "1", the rest of values will be "0".

Other control variables will be included are "female", "age" to see if
there's any effect to influence alcohol drinking frequency.

```{r}
#| label: tbl-1
library(ggplot2)

ggplot(ess11_1, aes(x = income_level, fill = alc_level)) +
  geom_bar() +
  labs(title = "Alcohol Consumption by Income Level",
       x = "Income",
       y = "Count",
       fill = "Alcohol Level",
       caption = "Data from European Social Survey Round11"
  )
```

## Descriptive statistics

As seen from @tbl-desc, the total sample size is 1251. Among 3 different
drink frequency groups, 347 people almost don't like to drink
frequently, 445 people drink in a low frequency, and 459 people drink
oftenly. The average age of respondents are 52.3 years old and the group
of frequent drinker has a bigger value of average age, 56.6. The average
income from the sample is located in the 6.3.

```{r}
#| label: tbl-desc
#| tbl-cap: Descriptive statistics for 2023 Finnish drinking patterns by income
library(gtsummary)

# for N & Percent
tbl_top <- ess11_1 |> 
  ungroup() |> 
  select(alc_level) |> 
  mutate(N = 1) |> 
  mutate(Pct = 100 * 1 / n()) |> 
  tbl_summary(by = alc_level, 
              type = list(N ~ 'continuous', Pct ~ 'continuous'),
              statistic = list(N ~ "{sum}", Pct ~ "{sum}%"),
              digits = list(N ~ 0, Pct ~ 1),
              label = list(Pct ~ " Share")) |> 
  add_overall(last = T) |> 
  modify_header(update = all_stat_cols() ~ "{level}") |> 
  modify_header(label = "Variable") |> 
  modify_footnote(update = everything() ~ NA)

tbl_bot <- ess11_1 |> 
  select(female, age, happy, income, alc_level) |> 
  tbl_summary(
    by = alc_level,
    statistic = list(all_continuous() ~ '{mean};({sd})',
      all_categorical() ~ '{p}%'),
    digits = list(
      all_categorical() ~ 1,
      all_continuous() ~ 1
    ),
    label = list(
      female ~ "Female (%)",          
      age ~ "Age (Mean; SD)",         
      happy ~ "Happy (%)",            
      income ~ "Income (Mean; SD)"    
    )
  ) |> 
  add_overall(col_label = "**Overall**", last = TRUE) |> 
  modify_header(update = all_stat_cols() ~ "{level}") |> 
  modify_header(label = "Variable") |> 
  modify_footnote(update = all_stat_cols() ~ NA)

#tab_spanner() 来控制表格列的显示
tbl_stack(list(tbl_top, tbl_bot)) |> 
  as_gt() |> 
  tab_spanner(label = "Alcohol level", columns = c("stat_1", "stat_2","stat_3"), id = 'Ch') |> 
  tab_source_note(
  source_note = md("*Source: European Social Survey Round11*")
) 
```

# Operationalized method

## Method1: Multinomial Logit Regression for H1

**Hypothesis 1.1**:Higher income people will have higher drinking
frequency, but it can be moderated by the level of happiness.

**Hypothesis 1.2**: Alcohol consumption from females is lower than
males, in the condition of same income level.

From hypotheses above, the analysis focus will be the relation between
income level and drinking frequency, and the sense of "happy" will be
interaction effect. Drinking frequency, as the outcome variable,
includes 3 levels so that multinomial logit regression is chosen for
modeling.

```{r}
#| include: false 
#| echo: false    
#| message: false  

library(nnet)
library(modelsummary)

#model
modela <- multinom(alc_level ~ income + female, data = ess11_1) #income & gender
modelb <- multinom(alc_level ~ income  +happy+income:happy + female, data = ess11_1) #happy effect

models <- list("ModelA" = modela,
               "ModelB" = modelb)
```

```{r}
#| label: tbl-3
modelsummary(models, shape =  term ~ model + response,
             fmt = 2,
             statistic =  c("conf.int", "std.error"),
             exponentiate = TRUE,stars = TRUE,
gof_map = c("nobs", "r.squared", "adj.r.squared", "AIC","BIC"),
             notes = list("Note: Odds ratios are reported with 95% confidence intervals.Data from European Social Survey Round11"),
 output = "gt") |> 
  gt::tab_header(
    title = "Multinomial Logit Regression Model Results"
  )

```

Both Model A and Model B shows "female" is less likely to drink more
when comparing to "male".Female have around 27% lower odds to drink at
low frequency, and 64% lower odds to drink at high frequency comparing
to male. This supports H1.2.

In Model A, higher income people tend to drink. Comparing to "Almost no"
drinker, "Low Alcohol" group has 11% higher odds per unit increase in
income and "Frequent Alcohol" group has 19% higher odds per unit
increase in income. However, when including "happy" in Model B, income
will be not that significant and it is a bit significant for frequent
drinkers. People who feel happy have 69% lower odds to have
low-frequency of drinking , comparing to the "Almost no" drinker. A
one-unit increase in income increases the odds of low-frequency drinking
by 20%, while the effect on frequent drinkers is positive effect but not
significant. Model B shows higher R2 and lower AIC result, so Model B
fits better. Thus, H1.1 is supported partially.

## Method2: Binary Logit Regression for H2

**Hypothesis 2**: High-income people are more likely to drink on
weekdays, comparing with low & middle-income people.

Binary logit regression model will be used to see if high-income people
tend to drink more on weekdays or not. "alcwkdy" in ESS refer to the
gram amount of drinking last time during weekday."alcwkdy" will be the
outcome varaible and set as a dummy variable, to see if specific income
group of people drink or not, "0=No" and "1=Yes". To avoid the impact of
alcohol drinking frequency, the interaction effect betweeen "income
level" and "alcohol level" will also be included in the model.

Results in @tbl-3 show that female are still significantly less likely
to drink on weekdays than male in any income-level groups. There's no
significance to prove that income level and their weekday drinking
behavior."High" income group shows a positive effect(OR:1.320) but it’s
not significant. However, the frequency of alcohol consumption matters.
The interaction results show that alcohol consumption level
significantly affects the relationship between income and weekday
drinking.For example,"Low" income group with "Low" alcohol consumption
have 3.392 times the odds of weekday drinking compared to the same group
with "Almost No" alcohol consumption

H2 doesn't get supported, suggesting that workplace demands can't be
proved to drive more drinking on weekday. Comparing with income
difference, the frequency of alcohol consumption have a great impact on
people who choose to drink on weekdays.

```{r}
ess11_wkday <- ess11_1 |> 
  mutate(choice = as.integer(alcwkdy > 0)) |> 
  select(idno, choice, income, female, happy, alc_level, income_level, age, alcfreq) |> 
  drop_na()

ess11_wkday <- ess11_wkday |> 
  mutate(income_level = relevel(income_level, ref = "Low Income"),
         alcohol_level = relevel(alc_level, ref = "Almost No"))

model_1 <- glm(choice ~ income_level + female + age ,
               family = binomial(link = "logit"),
               data = ess11_wkday)

model_2 <- glm(choice ~ income_level + female + age + income_level:alc_level,
               family = binomial(link = "logit"),
               data = ess11_wkday)

models2 <- list("Model 1" = model_1, "Model 2" = model_2)

```

```{r}

library(modelsummary)

modelsummary(
  models2,
  exponentiate = TRUE,
  stars = TRUE,
  output = "kableExtra",
  title = "Binary Logistic Regression Results for Weekday Drinking",
  notes = c(
    "Note: Estimates are odds ratios (OR).",
    "Data from European Social Survey Round 11."
  ),
  coef_map = c(
    "income_levelMiddle Income" = "Middle Income",
    "income_levelHigh Income" = "High Income",
    "female" = "Female (ref: Male)",
    "age" = "Age",
    "income_levelLow:alc_levelLow" = "Low Income × Low",
    "income_levelMiddle Income:alc_levelLow" = "Middle Income × Low",
    "income_levelHigh Income:alc_levelLow" = "High Income × Low",
    "income_levelLow:alc_levelFrequent" = "Low Income × Frequent",
    "income_levelMiddle Income:alc_levelFrequent" = "Middle Income × Frequent",
    "income_levelHigh Income:alc_levelFrequent" = "High Income × Frequent"
  )
) |> 
  kableExtra::kable_styling(full_width = FALSE)

```

### Model fit result

Comparing with Model 1 and Model 2 from H2, Model 2 has a higher
Log-Likelihood value and lower AIC, so it's a better fit in @tbl-5.

```{r}
#| label: tbl-5
library(lmtest)
library(DescTools) 

modelfit <- tibble(
  Model = c("Model A", "Model B"),
  #Evalution = c("0.62","0.63","0.67"),
  `Log-Likelihood` = c(round(logLik(models2[[1]]), 3), round(logLik(models2[[2]]), 3)),
  `AIC` = c(round(AIC(models2[[1]]), 3), round(AIC(models2[[2]]), 3))
)

modelfit |> 
  gt() |> 
  tab_header(
    title = "Model Fit Results for Models(H2)"
  )  
#  cols_width(px(100))
```

### Predictions for Model 2 in H2

Based on model fit results from Model 2, I will predict probability of
weekday drinking by different income and drinking frequency groups,
shown as @fig-1. Female are less likely to drink alcohol than male in
all income level groups. The higher drinking frequency, the higher
probability to drink on weekdays.Generally "Low" and "High" income
groups are more likely to drink on weekdays. In the "Almost No" and
"Frequent" drinking groups, "High" income group has more possibility to
drink more on weekday than the other 2 lower income groups. The "Middle"
income group is at the position of the lowest probability.

```{r}
#| label: fig-1
library(ggplot2)

grid_2 <- expand.grid(
  income_level = factor(c("Low Income", "Middle Income", "High Income"),
                        levels = c("Low Income", "Middle Income", "High Income")),
  female = c(0, 1),
  age = 55,
  alc_level = factor(c("Almost No", "Low", "Frequent"))
)

pred_grid <- predict(model_2, newdata = grid_2, type = "response")|> 
  as_tibble()  |>
  bind_cols(grid_2)
```

```{r}
#| echo: false
ggplot(pred_grid, aes(x = income_level, y = value, color = factor(female))) +
  geom_point(size = 3) +
  geom_line(aes(group = female)) +
  facet_wrap(~ alc_level, ncol = 3) +
  labs(
    title = "Weekday Drinking Probability by Income and Drinking Frequency",
    x = "Income Level",
    y = "Probability of Weekday Drinking",
    color = "Gender",
    caption = "Predicted probability based on ESS Round 11"
  ) +
  scale_color_manual(
    values = c("0" = "blue", "1" = "pink"),
    labels = c("Male", "Female")  # Change 0 and 1 to "Male" and "Female"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5),
    legend.position = "top"
  )
```


## Method 3: K-means clustering to find out different groups' behavior patterns

To explore whether distinct clusters of drinking behaviors emerge among
Finnish respondents, K-means clustering is applied to the ESS11 data.
This method groups individuals based on similarities in income, drinking
frequency (`alc_level`), gender (`female`), happiness (`happy`), and
weekday drinking (`choice`). The goal is to identify natural segments
that may align with or extend the findings from the regression models.

The number of clusters (k) is set to 3 to reflect the three income
levels and drinking frequency categories, though this can be adjusted
based on elbow method results. Variables are standardized to ensure
equal weighting.

```{r}
#| label: tbl-cluster-summary
#| tbl-cap: Summary Statistics by Cluster(K-means)

library(cluster)
#install.packages("factoextra")
library(factoextra)


# Prepare data for clustering
cluster_data <- ess11_wkday |> 
  select(income, female, happy, choice, alcfreq) |> 
  mutate(across(everything(), scale))  # Standardize variables

# Perform K-means clustering
set.seed(123)  # For reproducibility
kmeans_result <- kmeans(cluster_data, centers = 3, nstart = 25)

# Add cluster assignments to the original data
ess11_wkday$cluster <- kmeans_result$cluster

# Visualize clusters
fviz_cluster(kmeans_result, data = cluster_data, 
             geom = "point", 
             ellipse.type = "convex", 
             ggtheme = theme_minimal(),
             main = "K-means Clustering of Drinking Behaviors in different income-level") +
  labs(caption = "Data from European Social Survey Round 11")




```

The K-means clustering graph in @tbl-cluster-summary reveals three distinct groups:

\- **Cluster 1**: it shows high-income, frequent drinkers with higher
weekday drinking probability, predominantly male.

\- **Cluster 2**: it shows middle-income individuals with low drinking
frequency and minimal weekday drinking.

\- **Cluster 3**: it represents low-income, happier individuals with
moderate drinking habits.

These clusters provide additional context to the regression findings,
suggesting that income and drinking frequency interact with gender and
happiness to form behavioral segments. This supports the idea that
drinking in Finland may be socially driven rather than purely
income-based.


# Conclusion & Discussion

Hypothesis 1.1 and Hypothesis 1.2 get supported. H2 doesn't get
supported, but this analysis provides valuable insights of Finnish
drinking patterns for business or social alcohol management to make
strategies.

**Income vs Drinking frequency**:Higher income drives with higher
drinking frequency, but not necessarily on weekdays. And the drinking
frequency can be moderated by the level of happiness feeling.

**Male vs Female**: As for alcohol consumption, female's consumption is
lower than males in all the same income level groups and also less
amount of weekday drinking.

**Income vs Weekday drinking**: Weekday drinking is mostly driven by
frequent drinkers. Among 3 income-levels,middle-income people drink less
frequently.

**Business drinking vs Person drinking**: Drinking in Finland may be
more social and personal.

For the limitation of the analysis, the variable of "income" extracted
from ESS11 is ‘hinctnta’, a decile-based household income measure, may
differentiate individual income variations within households and lacks
precision.

"alcfreq" and "happy" are also from self report, so it can be different
in reality. To increase the precision, "income" variable in future
research can be conducted by individual level instead of household
level.

"alcqnt" only includes the amount respondents drink last time and the
sampling is small. It's better to have more records of their avaerge
amount drinks for weekday and weekend.

Age does not have an impact on drinking frequency or weekly drinking
amount, but this can be the sample selection since the average
respondents in each group are around 50 years old.

Causal effect between "Income" and "Happy" can be not clear. Higher
income people can be happier because they have more resouces, so IV can
be included in the next research.

**Bibioligraphy**

Bourdieu, P. (1984) Distinction: A Social Critique of the Judgement of
Taste. Routledge, pp. 5 & 41. Pareto, V. (1896). Cours d’économie
politique. Lausanne: F. Rouge.
